\textbf{COMP0085 Summative Assignment}

### 2023-01-03 submission.tex
add comments
split out ard code
edit report and rename variables for readibility

\begin{document}
\maketitle
\section*{Question 1}

\subsection*{(a)} The directed acyclic graph:

@@

%\begin{center}
%\begin{tikzpicture}[node distance={25mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
%    \node[main] (ABC) {$ABC$};
%    \node[main] (BCDF) [below of=ABC] {$BCDF$};
%    \node[main] (CFH) [below of=BCDF] {$CFH$};
%    \node[main] (FHI) [below left of=CFH] {$FHI$};
%    \node[main] (CEH) [below right of=CFH] {$CEH$};
%    \node[main] (EGH) [below of=CEH] {$EGH$};
%    \draw (ABC) -- (BCDF);
%    \draw (BCDF) -- (CFH);
%    \draw (CFH) -- (FHI);
%    \draw (CFH)  -- (CEH);
%    \draw (EGH)  -- (CEH);
%\end{tikzpicture}
%\end{center}

@@

The set $\{A, D, E, F, H\}$ is a non-unique smallest set of molecules such that if the concentrations of the species within the set are known, the concentrations of the others $\{B, C, G, I\}$ would all be independent (conditioned on the measured ones).

@@

where $\delta[\textbf{x}]$ are the concentration perturbations, $\epsilon \sim \mathcal{N} (0, \Psi)$, and the latent factors $z \sim \mathcal{N} (0, I)$.

@@

We can see that these simplify to the equations:

\begin{gather*}
    \delta[B] = \Lambda_{BA}z_A + \epsilon_B\\
    \delta[D] = \Lambda_{DC}z_C  + \epsilon_D\\
    \delta[E] = \Lambda_{EC}z_C  + \epsilon_E\\
    \delta[G] = \epsilon_G\\
\end{gather*}

Thus, we see that the only latent variables present are $z_A$ and $z_C$, so would expect to recover the factors of $A$ and $C$, the two parent nodes of the observations.

@@

\newpage
\section*{Question 2}

@@

Our distribution for $\textbf{w}$:
\[P(\textbf{w}) = \mathcal{N} \left(
\begin{bmatrix}
                 \mu_a \\
                 \mu_b
         \end{bmatrix} ,
\begin{bmatrix}
                 \sigma_a^2 & 0 \\
                 0 & \sigma_b^2
         \end{bmatrix}
\right)
  = \mathcal{N}(\mu_{\textbf{w}}, \Sigma_{\textbf{w}})
\]

Moreover, for our data $\mathcal{D} = \{\textbf{X}, \textbf{Y}\}$:
\[P(\mathcal{D} | \textbf{w}) = \mathcal{N} \left( \textbf{Y} - \textbf{w}^T \textbf{X}, \sigma^2 \textbf{I}
\right)
\]

where $\textbf{X} =  \begin{bmatrix}
                 t_1 & t_2 \cdots t_N \\
                 1 & 1 \cdots 1
         \end{bmatrix} \in  \mathbb{R}^{2 \times N}$ and $\textbf{Y} \in \mathbb{R}^{1 \times N}$.

@@

we can substitute the above distributions:
\[P(\textbf{w} | \mathcal{D}) \propto
  \exp \left(\frac{-1}{2 \sigma^2} \left( \textbf{Y} - \textbf{w}^T \textbf{X}\right) \left( \textbf{Y} - \textbf{w}^T \textbf{X}\right)^T \right)
\exp \left(\frac{-1}{2} \left( \textbf{w} -\mu_{\textbf{w}}\right)^T \Sigma_{\textbf{w}}^{-1} \left( \textbf{w} -\mu_{\textbf{w}}\right)\right)
\]

expanding:
\[\log P(\textbf{w} | \mathcal{D}) \propto
  \frac{-1}{2} \left( \frac{\textbf{Y} \textbf{Y}^T}{\sigma^2} - 2\textbf{w}^T \frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \textbf{w}^T \frac{\textbf{X} \textbf{X}^T}{\sigma^2} \textbf{w} +  \textbf{w}^T \Sigma_{\textbf{w}}^{-1}  \textbf{w} - 2\textbf{w}^T \Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} + \mu_{\textbf{w}}^T\Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} \right)
\]

collecting $\textbf{w}$ terms:
\[\log P(\textbf{w} | \mathcal{D}) \propto
  \frac{-1}{2} \left( \textbf{w}^T \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)  \textbf{w} - 2\textbf{w}^T \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right)  \right)
\]

Knowing that the posterior $P(\textbf{w} | \mathcal{D})$ will be Gaussian with mean $\bar{\mu}_w$ and covariance $\bar{\Sigma}_w$, we can see that expanding the exponent component would have the form:
\[\left( \textbf{w} - \bar{\mu}_w \right)^T \bar{\Sigma}_w^{-1} \left( \textbf{w} - \bar{\mu}_w \right) = \textbf{w}^T \bar{\Sigma}_w^{-1} \textbf{w} -2 \textbf{w}^T \bar{\Sigma}_w^{-1} \bar{\mu}_w + \bar{\mu}_w^T \bar{\Sigma}_w^{-1} \bar{\mu}_w\]

@@

and the posterior mean:
\[\bar{\mu}_w = \bar{\Sigma}_w \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right) \]

@@

Plotting the residuals:

@@

We can see that the residuals do not perfectly conform to our prior over $e(t) \sim \mathcal{N}(0, 1)$.
The density estimation shows that a mean of zero is a reasonable prior belief however the data does not seem to exhibit unit variance.
Also we know it's not iid because timeseries.

@@

We can make qualitative observations this kernel by visualising the covariance (gram) matrix:

@@

We can observe a striped pattern which indicate higher covariance at regular intervals.
This can be attributed to the sinusoidal term in the kernel and encourages sinusoidal functions.
Additionally, we can see that covariance values also decay as they are further away from the diagonal.
This can be attributed to the exponential term in the kernel, encouraging points closer in time to be more correlated and vice versa.
From our $CO_2$ data, we would want a class of functions which exhibit both of these behaviours as the data looks sinusoidal (seasonal with respect to each year) and correlations locally.

@@

\begin{itemize}
\item [$\theta$:] As $\theta$ increases, we see more pronounced periodic behavior in the sample function.
    The covariance matrix shows how increasing $\theta$ visually reveals the striped periodic component.
        This is expected because it is the parameter that adjusts the weight of the periodic component.
\item [$\sigma$:] As $\sigma$ increases, we see reduced periodic behaviour in the sample function.
    The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
    This is expected because it adjusts the lengthscale of the periodic portion of the kernel, which ends up dominating the function.
\item [$\phi$:] As $\phi$ increases, we see the ratio of the amplitude of the periodicity component of the sample function reduces compared to the baseline.
    The covariance matrix shows how increasing $\phi$ will start to increase the non-periodic component.
    This is expected because it adjusts the weight of the non-periodic portion of the kernel, thus the periodic component remains the same (i.e.same amplitude) but the large baseline shifts from increasing $\phi$ ends up dominating the function visually.
\item [$\eta$:] As $\eta$ increases we see smoother sample functions.
This is expected because the $\eta$ increases the lengthscale of the non-periodic component, allowing for smoother functions.
This causes the off-diagonals of the gram matrix to increase, however the periodic component is still maintained because $\eta$ doesn't affect the relative weight of the two components.
\item [$\tau$:] As $\tau$ increases, the period of the periodic function increases.
We can see this reflected in the stripes in the gram matrix getting further apart.
This makes sense because we are adjusting the period in the sinusoid function of the periodic term with $\tau$.
\item [$\zeta$:] As $\zeta$ increases, the function becomes less smooth.
    This is because the $\zeta$ parameter adjusts the weight of the $\delta_{s=t}$ parameter.
    This places stronger emphasis on the independence of each timestep, which can be seen with the reduction of relative magnitude of off-diagonals in the gram matrix.
    However, this is simply masking the periodic and squared-exponential terms as we can see with the increased magnitude of the functions as $\zeta$ increases.
\end{itemize}
\newpage
\subsection*{(e)}
Suitable values for hyper-parameters can be chosen through a combination of visual inspection and prior knowledge.
For example, it is a reasonable assumption that the $CO_2$ concentration levels have a strong yearly seasonality behaviour due to the cyclic changes in temperature, humidity, etc.
Thus we can choose $\tau=1$ to ensure functions with a period of one year to reflect this knowledge.
It can be difficult to quantitatively choose values for the other parameters as they can relate to the uncertainty exhibited in the data (i.e.the smoothness of the function).
One approach is to maximise:

\[\log P(\textbf{Y} | \textbf{X}) = -\frac{1}{2} \textbf{Y}^T (\textbf{K} + \sigma^2\textbf{I})^{-1} \textbf{Y} - \frac{1}{2} \log |\textbf{K}+ \sigma^2\textbf{I}| - \frac{n}{2}\log(2\pi)\]

the log-likelihood  of the posterior distribution with respect to the given data where $\textbf{K}$ is the gram matrix for the kernel (equation 2.30 from http://gaussianprocess.org/gpml/chapters/RW2.pdf).
We can define a loss function as the negative log-likelihood and employ gradient-based algorithms to find optimal parameters.

@@

We can analyse some of the changes in these parameters after training to gain some insights.
We can see that $\tau$ remains the same as we would expect given the yearly seasonality we have prior knowledge of.
On the other hand, the value for $\zeta$ is significantly reduced signifying that $\delta_{s=t}$ is not a very good kernel for representing the data as datapoints at different timesteps do exhibit correlations.

\newpage
\subsection*{(f)}
Extrapolating the $CO_2$ concentration levels:

@@

We can see that the extrapolation shows a continued increase in $CO_2$ in the future.
This follows our expectations given that the levels has been steadily increasing in the past.
Moreover, the concentration continues to exhibit yearly seasonality (for the trained extrapolation) as we would expect.
We can see that the conclusions can be quite sensitive to kernel hyperparameters when comparing the values from before and after training.
Prior to training, the extrapolated prediction is not representative of the given data, with pretty much no seasonal behaviour and very large uncertainty.
After training, we can see that the prediction is much more reasonable, and qualitatively the uncertainty bounds seem to exhibit the historical variability in the data.

@@

This procedure is not fully Bayesian because despite using a posterior estimate of our linear regression terms, we only use a point estimate when making prediction.
For a fully Bayesian approach, we should also incorporate the uncertainty of the linear regression parameters into our extrapolation/uncertainty bounds.
For our procedure, we only include the uncertainty of $g(t)$ however it can be observed in the plots that the trend is not perfectly linear so this should be reflected in the uncertainty of our extrapolation.
Another approach could be to add a linear kernel to our combined kernel function and model $f(t)$ directly with our kernel, removing the linear regression component in our procedure.
Thus our kernel extrapolation would incorporate the uncertainty of all components of our signal.

@@

\newpage
\section*{Question 3}

@@

The free energy is can be calculated as:
\[\mathcal{F}(q, \theta) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{q(\textbf{s})} + H[Q(\textbf{s})]\]

@@

Moreover, our mean field approximation:
\[q(\textbf{s}) = \prod_{i=1}^{K} q_{i} (s_i) \]

where $q_{i} (s_i) = \lambda_i^{s_i} (1-\lambda_i)^{(1-s_i)}$.

To compute the first term:

\[ P(\textbf{x} | \textbf{s}, \theta)  = \mathcal{N}\left( \sum_{i=1}^{K} s_i \mu_i, \sigma^2 \textbf{I} \right)}\]

substituting the appropriate terms:

\[ P(\textbf{x} | \textbf{s}, \theta)  = 2\pi^{-\frac{d}{2}} |\sigma^2 \textbf{I}|^{-\frac{1}{2}} \exp \left( -\frac{1}{2} \left(\textbf{x} - \sum_{i=1}^{K} s_i \mu_i\right)^T \frac{1}{\sigma^2} \textbf{I}  \left(\textbf{x} - \sum_{i=1}^{K} s_i \mu_i\right) \right) \]

@@

\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} s_i \mu_i   + \sum_{i=1}^{K} \sum_{i=1}^{K} s_i s_j \mu_i^T\mu_j \right) \]

@@

\[
\left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \left\langle s_i \right\rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \left\langle s_i s_j \right\rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]

@@

\[
\left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right)\]

where $\left\langle s_i s_i \right\rangle_{q_{i} (s_i)} = \left\langle s_i \right\rangle_{q_{i} (s_i)}$ because $s_i \in \{0, 1\}$.

To compute the second term:
\[ P(\textbf{s} | \theta) = \prod_{i=1}^{K}\pi_i^{s_i} (1-\pi_i)^{(1-s_i)}\]

@@

\[ \log P(\textbf{s} | \theta) = \sum_{i=1}^{K}s_i \log \pi_i + (1-s_i) \log (1-\pi_i)}\]

@@

\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \left\langle s_i \right\rangle_{q_{i} (s_i)} \log \pi_i + (1-\left\langle s_i\right\rangle_{q_{i} (s_i)}) \log (1-\pi_i)}\]

\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\]

@@

\[H\left[ q(\textbf{s})\right] = \sum_{i=1}^K H\left[ q_{i} (s_i)\right] \]

@@

\[H\left[ q(\textbf{s})\right] = - \sum_{i=1}^K \sum_{s_i \in \{0, 1\}} q_{i} (s_i) \log q_{i} (s_i) \]

@@

\[H\left[ q(\textbf{s})\right] = - \sum_{i=1}^K \lambda_i \log \lambda_i + (1-\lambda_i) \log (1-\lambda_i) \]

@@

\[
\begin{array}{l}
\tab \mathcal{F}(q, \theta) = \\
\tab \tab \frac{-d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right) \\
\tab \tab + \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\\
\tab \tab - \sum_{i=1}^K \lambda_i \log \lambda_i + (1-\lambda_i) \log (1-\lambda_i)
\end{array}
\]

To derive the partial update for $q_i(s_i)$ we take the variational derivative of the Lagrangian, enforcing the normalisation of $q_i$:

\[\frac{\partial}{\partial q_i}\left( \mathcal{F}(q, \theta) + \lambda^{LG} \int q_i -1)\right) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i) - 1 + \lambda^{LG}\]

@@

Setting this to zero we can solve for the $\lambda_i$ that maximises the free energy:

\[\log q_i(s_i) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} - 1 + \lambda^{LG}\]

@@

Thus, we can write the posterior:

@@

>>>>

\begin{document}
\maketitle
\section*{Question 1: A biochemical pathway}

\subsection*{(a)} We are given that:

\begin{itemize}
    \item The concentrations of both species B and C depend on that of A.
    \item Molecule C seems to redirect the process that produces enzyme D to produce enzyme E instead.
    \item D catalyses the production of F from B, while E catalyses the production of H from G.
    \item F and H then combine to form the end product I.
\end{itemize}

The corresponding directed acyclic graph:

@@

@@

The set $\{A, D, E, F, H\}$ is a non-unique smallest set of molecules such that if the concentrations of the species within the set are known, the concentrations of the others $\{B, C, G, I\}$ would all be independent (conditioned on the measured ones).
\newpage

@@

where $\delta[\textbf{x}]$ are the concentration perturbations, $\epsilon \sim \mathcal{N} (0, \Psi)$ are the reaction specific Gaussian noise variables, and $z \sim \mathcal{N} (0, I)$ are the latent factor, the parent concentration perturbations.

@@

Knowing \textbf{z} are the parent concentration perturbations, we can see that these simplify to a factor analysis set up:

\[
    \begin{bmatrix}
                 \delta[B]  \\
                 \delta[D]  \\
                 \delta[E]  \\
         \end{bmatrix} = \begin{bmatrix}
                 \Lambda_{BA} & 0   \\
                 0 & \Lambda_{DC} \\
                 0 & \Lambda_{EC} \\
         \end{bmatrix} \begin{bmatrix}
                 \delta[A]  \\
                 \delta[C]  \\
         \end{bmatrix} +  \begin{bmatrix}
                 \epsilon_B  \\
                 \epsilon_D  \\
                 \epsilon_E  \\
         \end{bmatrix}
\]

Thus, we would expect to recover the factors of $A$ and $C$, $\delta[A]$ and $\delta[C]$, the two parent nodes of the observations.
\newpage

@@

From factor analysis, we would recover $\delta[A]$ and $\delta[C]$.
In theory, if these concentration perturbations had unknown parent factors, we could use the results to recover the concentration perturbations of those parent features.
However, in our DAG this is not hte case, so we would not be able to recover any other species in the cascade because  they are all downstream from the known concentration perturbations and those recovered from factor analysis.

\newpage
\section*{Question 2: Bayesian linear and Gaussian process regression}

@@

Our distribution for $\textbf{w}$:
\[P(\textbf{w}) = \mathcal{N}(\mu_{\textbf{w}}, \Sigma_{\textbf{w}}) = \mathcal{N} \left(
\begin{bmatrix}
                 \mu_a \\
                 \mu_b
         \end{bmatrix} ,
\begin{bmatrix}
                 \sigma_a^2 & 0 \\
                 0 & \sigma_b^2
         \end{bmatrix}
\right)

\]

For our data $\mathcal{D} = \{\textbf{X}, \textbf{y}\}$:
\[P(\mathcal{D} | \textbf{w}) = \prod_{t=1}^{T}\mathcal{N} \left( y_t - \textbf{w}^T \textbf{x}_t, \sigma^2
\right)
\]

where $\textbf{x}_t =  \begin{bmatrix}
                 t_1 \\
                 1
         \end{bmatrix} \in  \mathbb{R}^{2 \times 1}$ and $y_t \in \mathbb{R}^{1 \times 1}$ our response $f_{obs}(t_1)$.

@@

we can substitute the above distributions:
\[P(\textbf{w} | \mathcal{D}) \propto
  \exp \left(\frac{-1}{2 \sigma^2} \left( \textbf{y} - \textbf{w}^T \textbf{X}\right) \left( \textbf{y} - \textbf{w}^T \textbf{X}\right)^T \right)
\exp \left(\frac{-1}{2} \left( \textbf{w} -\mu_{\textbf{w}}\right)^T \Sigma_{\textbf{w}}^{-1} \left( \textbf{w} -\mu_{\textbf{w}}\right)\right)
\]

where $\textbf{X} =  \begin{bmatrix}
                 t_1 & t_2 \cdots t_T \\
                 1 & 1 \cdots 1
         \end{bmatrix} \in  \mathbb{R}^{2 \times T}$ and $\textbf{y} \in \mathbb{R}^{1 \times T}$.
Expanding:

\[\log P(\textbf{w} | \mathcal{D}) \propto
  \frac{-1}{2} \left( \frac{\textbf{y} \textbf{y}^T}{\sigma^2} - 2\textbf{w}^T \frac{\textbf{X}\textbf{y}^T}{\sigma^2} + \textbf{w}^T \frac{\textbf{X} \textbf{X}^T}{\sigma^2} \textbf{w} +  \textbf{w}^T \Sigma_{\textbf{w}}^{-1}  \textbf{w} - 2\textbf{w}^T \Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} + \mu_{\textbf{w}}^T\Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} \right)
\]

collecting $\textbf{w}$ terms:
\[\log P(\textbf{w} | \mathcal{D}) \propto
  \frac{-1}{2} \left( \textbf{w}^T \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)  \textbf{w} - 2\textbf{w}^T \left(\frac{\textbf{X}\textbf{y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right)  \right)
\]

Knowing that the posterior $P(\textbf{w} | \mathcal{D})$ will be Gaussian with mean $\bar{\mu}_w$ and covariance $\bar{\Sigma}_w$, we can see that expanding the exponent component of a Gaussian would have the form:

\[\left( \textbf{w} - \bar{\mu}_w \right)^T \bar{\Sigma}_w^{-1} \left( \textbf{w} - \bar{\mu}_w \right) = \textbf{w}^T \bar{\Sigma}_w^{-1} \textbf{w} -2 \textbf{w}^T \bar{\Sigma}_w^{-1} \bar{\mu}_w + \bar{\mu}_w^T \bar{\Sigma}_w^{-1} \bar{\mu}_w\]

@@

and the posterior mean:
\[\bar{\mu}_w = \bar{\Sigma}_w \left(\frac{\textbf{X}\textbf{y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right) \]

@@

Plotting the residuals $g_{obs}(t) = f_{obs}(t) - (a_{MAP}t + b_{MAP})$:

@@

We can see that the residuals do not perfectly conform to our prior over $e(t) \sim \mathcal{N}(0, 1)$.
The density estimation shows that a mean of zero is a reasonable prior belief however the data does not seem to exhibit unit variance.
Also, observing $g_{obs}(t)$ we see that there is periodic structure in the timeseries meaning that the data is not independently and identically distributed as $e(t) \sim \mathcal{N}(0, 1)$ would suggest.
Otherwise, we would expect $g(t)$ to look like random noise.

@@

We can make qualitative observations of this kernel by visualising the covariance (gram) matrix:

@@

We can observe a striped pattern which indicate higher covariance at regular intervals.
This can be attributed to the sinusoidal term in the kernel and encourages periodic structure in functions.
Additionally, we can see that covariance values also decay as they are further away from the diagonal.
This can be attributed to the exponential term in the kernel, encouraging points closer in time to be more correlated and vice versa.
From visualising $g(t)$, we would want to model this with a class of functions which exhibit both of these behaviours as $g(t)$ looks periodic (seasonal with respect to each year) and locally correlated (smooth).

@@

\begin{itemize}
\item [$\theta$:] As $\theta$ increases, we see more pronounced periodic behavior in the sample function by increasing its amplitude.
    The covariance matrix shows how increasing $\theta$ visually reveals the striped periodic component.
        This is expected because it is the parameter that adjusts the amplitude of the periodic component.
\item [$\sigma$:] As $\sigma$ increases, we see smoother periodic behaviour in the sample function.
    The covariance matrix shows how increasing $\sigma$ will increase covariance values around each periodic stripe of the covariance matrix.
    This is expected because it adjusts the lengthscale of the periodic portion of the kernel.
\item [$\phi$:] As $\phi$ increases, we see the ratio of the amplitude of the periodicity component of the sample function reduces compared to the baseline.
    The covariance matrix shows how increasing $\phi$ will start to increase the non-periodic component masking the periodic stripes.
    This is expected because it adjusts the weight of the squared exponential portion of the kernel, thus the periodic component remains the same (i.e.same amplitude) but the large baseline shifts from increasing $\phi$ ends up dominating the function visually.
\item [$\eta$:] As $\eta$ increases we see smoother sample functions.
This is expected because the $\eta$ increases the lengthscale of the non-periodic component, allowing for smoother functions.
This causes the off-diagonals of the gram matrix to increase, however the periodic component is still maintained because $\eta$ doesn't affect the relative weighting between the two components.
\item [$\tau$:] As $\tau$ increases, the period of the periodic function increases.
We can see this reflected in the stripes in the gram matrix getting further apart and the period of the samples.
This makes sense because we are adjusting the period parameter in the sinusoid function of the periodic term with $\tau$.
\item [$\zeta$:] As $\zeta$ increases, the function becomes less smooth.
    This is because the $\zeta$ parameter adjusts the weight of the $\delta_{s=t}$ kernel.
    This places stronger emphasis on the independence of each timestep, which can be seen with the reduction of relative magnitude of off-diagonals in the gram matrix.
    This masks the periodic and squared-exponential components as we can see with the increased magnitude of the functions as $\zeta$ increases.
\end{itemize}
\subsection*{(e)}
Suitable values for hyper-parameters can be chosen through a combination of visual inspection and prior knowledge.
For example, it is a reasonable assumption that the $CO_2$ concentration levels have a strong yearly seasonality behaviour due to the cyclic changes in temperature, humidity, etc.
This behaviour cannot be modelled by our Bayesian linear regression and is reflected in our residuals $g(t)$.
Thus we can choose $\tau=1$ to ensure functions with a period of one year to reflect this prior knowledge.
It can be difficult to quantitatively choose values for the other parameters as they can relate to the uncertainty exhibited in the data (i.e.the smoothness of the function).
One approach is to maximise:

\[\log P(\textbf{y} | \textbf{X}) = -\frac{1}{2} \textbf{y}^T (\textbf{K} + \sigma^2\textbf{I})^{-1} \textbf{y} - \frac{1}{2} \log |\textbf{K}+ \sigma^2\textbf{I}| - \frac{n}{2}\log(2\pi)\]

the log-likelihood  of the posterior distribution with respect to the given data where $\textbf{K}$ is the gram matrix for the kernel (equation 2.30 from http://gaussianprocess.org/gpml/chapters/RW2.pdf).
We define a loss function as the negative log-likelihood and employ gradient-based algorithms to find optimal values for these parameters.

@@

We can analyse some of the changes in these parameters from optimisation, to gain some insights.
We can see that $\tau$ remains the same as we would expect, given the yearly seasonality we knew apriori.
On the other hand, the value for $\zeta$ is significantly reduced signifying that $\delta_{s=t}$ is not a very good kernel for representing the data as datapoints at different timesteps do exhibit correlations.

\newpage
\subsection*{(f)}
Extrapolating the $CO_2$ concentration levels using both the untrained and trained hyperparameters:

@@

We can see that the extrapolation shows a continued increase in $CO_2$ in the future.
This follows our expectations given that the levels has been steadily increasing in the past and our Bayesian linear regression model assumption.
Moreover, the concentration continues to exhibit yearly seasonality (for the trained extrapolation) as we would expect through the choice of our kernel for modelling the residuals $g(t)$.
We can see that the conclusions can be quite sensitive to kernel hyperparameters when comparing the extrapolations from before and after training.
Prior to training, the extrapolated prediction is not representative of the given data with very large uncertainty.
After training, we can see that the prediction is much more reasonable, and qualitatively the uncertainty bounds seem to exhibit the historical variability in the data.

@@

This procedure is not fully Bayesian because despite using a posterior estimate of our linear regression terms, we only use a MAP point estimate when making prediction.
For a fully Bayesian approach, we should also incorporate the uncertainty of the linear regression parameters into our extrapolation/uncertainty bounds.
For our procedure, we only include the uncertainty of $g(t)$ however it can be observed in the plots that the trend is not perfectly linear so this should be reflected in the uncertainty of our extrapolation.
Another approach could be to add a linear kernel to our combined kernel function and model $f(t)$ directly with our kernel, removing the linear regression component in our procedure.
Thus our kernel extrapolation would incorporate the uncertainty of all components of our signal.

@@

\newpage
\section*{Question 3: Mean-field learning}

@@

The free energy is can be expressed as:
\[\mathcal{F}(q, \theta) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{q(\textbf{s})} + H[Q(\textbf{s})]\]

@@

Moreover, our mean field approximation is:
\[q(\textbf{s}) = \prod_{k=1}^{K} q_{k} (s_k) \]

where $q_{k} (s_k) = (\lambda_k)^{s_k} (1-\lambda_k)^{(1-s_k)}$.

Given that:

\[ P(\textbf{x} | \textbf{s}, \theta)  = \mathcal{N}\left( \sum_{k=1}^{K} s_k \mu_k, \sigma^2 \textbf{I} \right)}\]

we can substitute the appropriate terms:

\[ P(\textbf{x} | \textbf{s}, \theta)  = 2\pi^{-\frac{d}{2}} |\sigma^2 \textbf{I}|^{-\frac{1}{2}} \exp \left( -\frac{1}{2} \left(\textbf{x} - \sum_{k=1}^{K} s_k \mu_k\right)^T \frac{1}{\sigma^2} \textbf{I}  \left(\textbf{x} - \sum_{k=1}^{K} s_k \mu_k\right) \right) \]

@@

\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} s_k \mu_k   + \sum_{k=1}^{K} \sum_{k=1}^{K} s_k s_{k'} \mu_k^T\mu_{k'} \right) \]

@@

\[
\left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \left\langle s_k \right\rangle_{q_{k} (s_k)} \mu_k   + \sum_{k=1}^{K} \sum_{k'=1}^{K} \left\langle s_k s_{k'} \right\rangle_{q_{k} (s_k) q_{k\} (s_{k'})} \mu_k^T\mu_{k'} \right)\]

@@

\[
\left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K}  \lambda_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  \lambda_k \lambda_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  \lambda_k \mu_k^T\mu_k \right)\]

where $\left\langle s_k s_k \right\rangle_{q_{k} (s_k)} = \left\langle s_k \right\rangle_{q_{k} (s_k)}$ because $s_k \in \{0, 1\}$.

Given that:

\[ P(\textbf{s} | \theta) = \prod_{k=1}^{K}\pi_k^{s_k} (1-\pi_k)^{(1-s_k)}\]

@@

\[ \log P(\textbf{s} | \theta) = \sum_{k=1}^{K}s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\]

@@

\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{k=1}^{K} \left\langle s_k \right\rangle_{q_{k} (s_k)} \log \pi_k + (1-\left\langle s_k\right\rangle_{q_{k} (s_k)}) \log (1-\pi_k)}\]

Evaluating the expectations with respect to $q(\textbf{s})$:

\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{k=1}^{K} \lambda_k \log \pi_k + (1-\lambda_k) \log (1-\pi_k)}\]

@@

\[H\left[ q(\textbf{s})\right] = \sum_{k=1}^K H\left[ q_{k} (s_k)\right] \]

@@

\[H\left[ q(\textbf{s})\right] = - \sum_{k=1}^K \sum_{s_k \in \{0, 1\}} q_{k} (s_k) \log q_{k} (s_k) \]

@@

\[H\left[ q(\textbf{s})\right] = - \sum_{k=1}^K \lambda_k \log \lambda_k + (1-\lambda_k) \log (1-\lambda_k) \]

@@

\[
\begin{array}{l}
\tab \mathcal{F}(q, \theta) = \\
\tab \tab \frac{-d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K}  \lambda_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  \lambda_k \lambda_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  \lambda_k \mu_k^T\mu_k \right) \\
\tab \tab + \sum_{k=1}^{K} \lambda_k \log \pi_k + (1-\lambda_k) \log (1-\pi_k)}\\
\tab \tab - \sum_{k=1}^K \lambda_k \log \lambda_k + (1-\lambda_k) \log (1-\lambda_k)
\end{array}
\]

To derive the partial update for $q_k(s_k)$ we take the variational derivative of the Lagrangian, enforcing the normalisation of $q_k$:

\[\frac{\partial}{\partial q_k}\left( \mathcal{F}(q, \theta) + \lambda^{LG} \int q_k -1)\right) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} - \log q_k(s_k) - 1 + \lambda^{LG}\]

@@

Setting this to zero we can solve for the $\lambda_k$ that maximises the free energy:

\[\log q_k(s_k) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} - 1 + \lambda^{LG}\]

@@

Thus, we can express the posterior as:

@@

We are left with:
